---
title: "Quartely report of swerb data"
author: "David Jansen"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lastpage}
output:
  pdf_document: default
keep_tex: yes
classoption: landscape
---

<!--- Define Headers and Footers --->
\fancypagestyle{plain}{%
  \renewcommand{\headrulewidth}{0pt}%
  \fancyhf{}%
  \fancyfoot[R]{\footnotesize Page \thepage\, of\, \pageref*{LastPage}}
  \setlength\footskip{0pt}
}

\pagestyle{plain}

```{r, include=FALSE}
end_of_quarter <- lubridate::ymd('2022-12-31')
```

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE, 
                      warning = FALSE,
                      message = FALSE,
#                     fig.fullwidth = TRUE,
                      fig.height = 8, fig.width = 11,
                      include =TRUE)
knitr::opts_knit$set(root.dir = "..")
```

This is the report for the `r scales::ordinal(lubridate::quarter(end_of_quarter))` quarter `r lubridate::year(end_of_quarter)`.

\tableofcontents

```{r, include = FALSE}
list.of.packages <- list("lubridate", "RPostgreSQL", "tidyverse", "gridExtra", "grid", "ggplot2", "rstudioapi")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if (length(new.packages)) install.packages(unlist(new.packages))
lapply(list.of.packages, require, character.only = T)
```

# Swerb data
```{r}
## The next part sets the colors and other aesthetics

observer_colors <- c("ILS" = "#4b6cf2",
                     "JKW" = "#FF3200", 
                     "RSM" = "#1BB6AF", 
                     #"SNS" = "#E9A17C", 
                     "multiple" = "#878787",
                     "no observers" = "#d8d6d6") 

observer_colors <- c("ILS" = "#4b6cf2",
                     "JKW" = "#FF3200", 
                     "RSM" = "#1BB6AF"
                     #,"SNS" = "#E9A17C"
                     ) 

observer_sizes <- c("ILS" = 5,
                     "JKW" = 5, 
                     "RSM" = 5, 
                     #"SNS" = 5, 
                     "multiple" = 3,
                     "no observers" = 3) 

adcode_colors <- c("Descent, before arrival" = "#878787", 
                   "MDT observed" = "#FF3200", 
                   "Descent, time unknown" = "#1BB6AF",
                   "MAT observed" = "#4b6cf2",
                   "Ascent, after departure" = "#d8d6d6") 

driver_colors <-   c(#"ILS" = "#4b6cf2",
                     "JKW" = "#FF3200", 
                     "BOO" = "#1BB6AF", 
                     "GYM" = "#E9A17C", 
                     "ALM" = "#f1f507",
                     "multiple" = "#878787",
                     "no drivers" = "#d8d6d6") 

driver_sizes <- c(#"ILS" = 5,
                     "JKW" = 5, 
                     "BOO" = 5, 
                     "GYM" = 5, 
                     "ALM" = 5, 
                     "multiple" = 3,
                     "no drivers" = 3) 
```

```{r}
## This parts gets the data
source("./R/get_data.R")
```

```{r}
## getting data ready with some time and date related mutations
## it also adds data on observers and drivers

swerb_data <-  
  swerb_data_raw %>% 
  arrange(date) %>% 
  filter(between(date, end_of_quarter - years(2) + days(1), end_of_quarter)) %>%  
  mutate(year_quarter = fct_inorder(paste(scales::ordinal(lubridate::quarter(date)), lubridate::year(date), sep="-"), ordered=TRUE))  %>% 
  mutate(time = if_else(is.na(time), depart_time, time)) %>% 
  mutate(trip = if_else(am(hms(time))==TRUE, "Morning", "Afternoon")) %>%
  #mutate(trip = factor(trip, levels = c("Morning", "Afternoon"))) %>%
  group_by(did) %>% 
  mutate(times_of_interest = if_else(event == 'B', min(time),
                                     if_else(event == 'E', max(time), NA_character_))) %>% 
  ungroup() %>% 
  filter(time == times_of_interest) %>%
  mutate(observer = ifelse(is.na(observers), "no observers",
                           if_else(nchar(observers) > 3, "multiple", 
                                   observers))) %>%
  
  mutate(driver = ifelse(is.na(drivers), "no drivers",
                         if_else(nchar(drivers) > 3, "multiple", 
                                 drivers))) %>%
  mutate(travel_time = case_when(event == 'B' ~ time_length(hms(time) - hms(depart_time),unit="minute"))) %>% 
  mutate(adcode_description = if_else(adcode_description == "Descent", "MDT observed",
                                    if_else(adcode_description == "Ascent", "MAT observed", 
                                            adcode_description))) %>% 
  mutate(observer = factor(observer, levels = c("ILS", "JKW", "RSM", "SNS", "multiple", "no observers"))) %>% 
  mutate(driver = factor(driver, levels = c("ALM", "BOO", "GYM", "JKW", "multiple", "no drivers"))) %>% 
  mutate(adcode_description = factor(adcode_description, levels = c("Descent, before arrival", "MDT observed", 
                                                                    "Descent, time unknown",
                                                                    "MAT observed", "Ascent, after departure"))) %>% 
  ## Filter on common observers and drivers                                                                    
  filter(observer %in% c("ILS", "JKW", "RSM", "SNS")) %>% 
  filter(driver %in% c("BOO", "GYM", "ILS", "JKW", "ALM")) %>%
  filter(!is.na(depart_time)) %>% 
  filter(!(trip == "Afternoon" & am(hms(depart_time)))) %>% 
  droplevels()

## calculate outliers in depart times
swerb_data_outlier <- swerb_data %>%
  left_join(swerb_data %>%
              group_by(trip) %>% 
              summarize(mean_depart_time =  mean(chron::times(depart_time), na.rm = TRUE),
                        sd_depart_time = sd(hms(depart_time), na.rm = TRUE))) %>% 
  filter(!((hms(depart_time) > hms(mean_depart_time) - 2 * seconds(sd_depart_time)) &
        (hms(depart_time) < hms(mean_depart_time) + 2 * seconds(sd_depart_time)))) %>% 
  mutate(depart_time = as.POSIXct(strptime(depart_time, format="%H:%M:%S")))


swerb_data_outlier <- swerb_data_outlier %>% 
  select(trip, year_quarter, observer, depart_time) %>% 
  arrange(year_quarter, depart_time) %>% 
  complete(observer, nesting(trip, year_quarter)) %>% 
    filter(observer %in% c("ILS", "JKW", "SNS", "RSM")) %>% 
  droplevels()
  
## gettings ome mean values for the table  
swerb_data <- swerb_data %>%
  left_join(swerb_data %>%
              group_by(trip) %>% 
              summarize(mean_depart_time =  mean(chron::times(depart_time), na.rm = TRUE),
                        sd_depart_time = sd(hms(depart_time), na.rm = TRUE))) %>% 
  filter(((hms(depart_time) > hms(mean_depart_time) - 2 * seconds(sd_depart_time)) &
        (hms(depart_time) < hms(mean_depart_time) + 2 * seconds(sd_depart_time)))) 

## get a summary of the quarterly data
swerb_summary_quarterly <- swerb_data %>%
  filter(!is.na(depart_time)) %>%
  group_by(observer, year_quarter, trip) %>%
    summarise(average_depart_time = mean(chron::times(depart_time), na.rm = TRUE),
              average_travel_time = mean(travel_time, na.rm = TRUE),
              sd_travel_time = sd(travel_time, na.rm = TRUE),
              trips = n(),
              missing_time = sum(is.na(depart_time))
              ) %>% 
  ungroup() %>% 
  mutate(average_depart_time = as.POSIXct(strptime(average_depart_time, format="%H:%M:%S"))) %>% 
  mutate(trips = if_else(observer %in% c("no observers", "multiple"), NA_integer_, trips))

## get a summary of the per observer
swerb_summary_observer <- swerb_data %>%
  filter(!is.na(depart_time)) %>%
  ## create filter to only get data for last year
  filter(between(date, left = end_of_quarter - years(1) + days(1), right = end_of_quarter)) %>% 
  group_by(observer, trip) %>%
  summarise(average_depart_time = mean(chron::times(depart_time), na.rm = TRUE),
            average_travel_time = mean(travel_time, na.rm = TRUE),
            sd_travel_time = sd(travel_time, na.rm = TRUE),
            trips = n(),
            missing_time = sum(is.na(depart_time))
              ) %>% 
  ungroup() %>% 
  mutate(average_depart_time = as.POSIXct(strptime(average_depart_time, format="%H:%M:%S"))) %>% 
  mutate(trips = if_else(observer %in% c("no observers", "multiple"), NA_integer_, trips))

swerb_summary_observer_yearly <- swerb_data %>%
  filter(!is.na(depart_time)) %>%
  ## create filter to only get data for last year
  filter(between(date, left = end_of_quarter - years(2) + days(1), right = end_of_quarter)) %>% 
  group_by(observer, trip) %>%
  summarise(average_depart_time = mean(chron::times(depart_time), na.rm = TRUE),
            average_travel_time = mean(travel_time, na.rm = TRUE),
            sd_travel_time = sd(travel_time, na.rm = TRUE),
            trips = n(),
            missing_time = sum(is.na(depart_time))
              ) %>% 
  ungroup() %>% 
  mutate(average_depart_time = as.POSIXct(strptime(average_depart_time, format="%H:%M:%S"))) %>% 
  mutate(trips = if_else(observer %in% c("no observers", "multiple"), NA_integer_, trips))

swerb_summary_observer_quarter <- swerb_data %>%
  filter(!is.na(depart_time)) %>%
  ## create filter to only get data for quarter
  filter(between(date, left = end_of_quarter %m-% months(3) %m+% days(1), right = end_of_quarter)) %>% 
  group_by(observer, trip) %>%
  summarise(average_depart_time = mean(chron::times(depart_time), na.rm = TRUE),
            average_travel_time = mean(travel_time, na.rm = TRUE),
            sd_travel_time = sd(travel_time, na.rm = TRUE),
            trips = n(),
            missing_time = sum(is.na(depart_time))
              ) %>% 
  ungroup() %>% 
  mutate(average_depart_time = as.POSIXct(strptime(average_depart_time, format="%H:%M:%S"))) %>% 
  mutate(trips = if_else(observer %in% c("no observers", "multiple"), NA_integer_, trips))

options(digits = 1)

## make the tables that appear in the plots
dt_am_table <- swerb_summary_observer_yearly %>%
  filter(trip=="Morning") %>% 
  mutate("average depart time\n last 2 years" = 
           format(average_depart_time, format='%H:%M')) %>%
  select(observer, "average depart time\n last 2 years") %>% 
  inner_join(swerb_summary_observer_quarter %>%
               filter(trip=="Morning") %>%
               mutate("average depart time\n this quarter" = 
               format(average_depart_time, format='%H:%M')) %>% 
               select(observer, "average depart time\n this quarter"))

dt_pm_table <- swerb_summary_observer_yearly %>%
  filter(trip=="Afternoon") %>% 
  mutate("average depart time\n last 2 years" = 
           format(average_depart_time, format='%H:%M')) %>%
  select(observer, "average depart time\n last 2 years") %>% 
  inner_join(swerb_summary_observer_quarter %>%
               filter(trip=="Afternoon") %>%
               mutate("average depart time\n this quarter" = 
               format(average_depart_time, format='%H:%M')) %>% 
               select(observer, "average depart time\n this quarter"))

tt_am_table <- swerb_summary_observer %>%
  filter(trip=="Morning") %>% 
  mutate("average travel time" = average_travel_time) %>% 
  select(observer, "average travel time") 

tt_pm_table <- swerb_summary_observer %>%
  filter(trip=="Afternoon") %>% 
  mutate("average travel time" = average_travel_time) %>% 
  select(observer, "average travel time") 

tt_table <- swerb_summary_observer %>%
  mutate("average travel time" = average_travel_time) %>% 
  select(observer, trip, "average travel time") %>% 
  spread(key = trip, value = "average travel time")

## some details to help with the tables
library(gridExtra)
tt2 <- ttheme_default(base_size = 7, base_colour = "black",
                      parse = FALSE, padding = unit(c(2, 2), "mm"))
```

## Depart times
```{r, depart_times_monthly}
dt <- swerb_summary_quarterly %>% 
  arrange(desc(observer)) %>% 
 filter(observer %in% c("ILS", "JKW", "SNS", "RSM")) %>% 
  mutate(y_min = if_else(trip == "Morning", 
                        as.POSIXct(strptime("5:00:00", format="%H:%M:%S")),
                        as.POSIXct(strptime("12:00:00", format="%H:%M:%S"))),
         y_max = if_else(trip == "Morning", 
                        as.POSIXct(strptime("5:45:00", format="%H:%M:%S")),
                        as.POSIXct(strptime("12:45:00", format="%H:%M:%S")))) %>%
  ggplot(aes(x=year_quarter, y=average_depart_time, color = observer, group = observer)) + 
    geom_blank(aes(y = y_min)) +
    geom_blank(aes(y = y_max)) +
    geom_point(aes(size = observer),position=position_dodge(width = .4)) +
    geom_point(data = swerb_data_outlier,  aes(y = depart_time),shape = 8,
               position=position_dodge(width = .4),
               show.legend = FALSE) +
    scale_size_manual(values = observer_sizes, guide = 'none') +
    geom_text(aes(label=trips), size = 2, color = "black", position=position_dodge(width = .4)) +
    facet_wrap(~trip, nrow=2, scales = "free_y") +
    scale_color_manual(values=observer_colors)  +
    annotation_custom(tableGrob(dt_am_table, theme = tt2),
                      xmin=6,
                      ymin=as.POSIXct(strptime("5:00:00", format="%H:%M:%S")), 
                      ymax = as.POSIXct(strptime("5:25:00", format="%H:%M:%S"))) +
    annotation_custom(tableGrob(dt_pm_table, theme = tt2),
                      xmin=6,
                      ymin=as.POSIXct(strptime("13:00:00", format="%H:%M:%S")), 
                      ymax = as.POSIXct(strptime("13:25:00", format="%H:%M:%S"))) +
    ggtitle("Average depart times from camp") +
    # xlab("Year-month") +
    ylab("Time") +
    ggthemes::theme_tufte(base_size = 12) +
    theme(axis.text.x = element_text(angle = 90)
          , legend.position = 'bottom'
          , panel.grid.major.y = element_line( size=.1, color="black" )
          , panel.background = element_rect(fill = NA, color = "black")
          )

dt + labs(caption="Table has the average values calculated over 2 running years as well as last quarter,\n dots are the mean for that quarter (numbers inside dots are the number of trips by that observer)\n stars are outliers (2SD away from mean)")
```


## Travel times
### Monthly
```{r, travel_time_monthly}
plot_tt <- swerb_summary_quarterly %>% 
  filter(!(trip == "Afternoon" & average_depart_time < as.POSIXct(strptime("12:00:00", format="%H:%M:%S")))) %>% 
  ggplot(aes(x=year_quarter, y=average_travel_time, color = observer)) + 
    geom_point(aes(size = observer), position=position_dodge(width = .4)) +
    geom_errorbar(aes(ymin=average_travel_time - sd_travel_time, 
                      ymax=average_travel_time + sd_travel_time), 
                  position=position_dodge(width = .4), size = 0.25) +
    scale_size_manual(values = observer_sizes, guide = 'none') +
    geom_text(aes(label=trips), size = 2, color = "black", position=position_dodge(width = .4)) +
    facet_wrap(~trip, nrow=2) +
    scale_y_continuous(breaks = seq(45,120,by = 15)) +
    scale_color_manual(values=observer_colors)  +
    # annotation_custom(tableGrob(tt_table, theme = tt2),
    #                     xmin=7,
    #                     ymax= 40) +
    ggtitle("Average travel times from camp to first group") +
    # xlab("Year-month") +
    ylab("Duration in minutes") +
    ggthemes::theme_tufte() +
    theme(axis.text.x = element_text(angle = 90)
          , legend.position = 'bottom'
          , panel.grid.major.y = element_line( size=.1, color="black" )
          , panel.background = element_rect(fill = NA, color = "black")
          )  +
  labs(caption="Table has the average values calculated over running years, dots are the mean for that quarter and numbers inside dots are the number of trips by that observer",
       color = "Observer",size = NA) 

plot_tt
  
```


## Ascent and descent times
### Observers
```{r}
brks <- c(0, 0.25, 0.5, 0.75, 1)

adcode_colors <- c("Descent, before arrival" = "#878787", 
                   "MDT observed" = "#FF3200", 
                   "Descent, time unknown" = "#1BB6AF",
                   "MAT observed" = "#4b6cf2",
                   "Ascent, after departure" = "#d8d6d6") 


swerb_data %>% 
  filter(nchar(observers) == 3) %>%
  filter(observer != "SNS") %>% 
  filter((trip == "Morning" & event == "B") |  trip == "Afternoon" & event == "E") %>%
  filter(!is.na(observers)) %>%
  group_by(observer, year_quarter, trip) %>%
  mutate(trips=n()) %>% 
  group_by(observer, year_quarter, trip, trips, adcode_description) %>%
  summarise(counts=n()) %>% 
  mutate(perc=counts/trips) %>%
  ggplot(aes(x = factor(year_quarter), y = perc, fill = factor(adcode_description))) +
    geom_bar(stat="identity", width = 0.7) +
    geom_text(aes(label=trips, factor(year_quarter), y = 0.1), size = 3, color = "black") +
    scale_y_continuous(breaks = brks, labels = scales::percent(brks)) +
    labs(x = "Year-month", y = "Percentages", fill = "Ascent and Descent codes") +
    scale_fill_manual(values=adcode_colors)  +
    facet_grid(trip~observer) +
    ggtitle("Proportion of adcodes") +
    ggthemes::theme_tufte() +
    theme(axis.text.x = element_text(angle = 90)
          , legend.position = 'bottom'
          , panel.grid.major.y = element_line( size=.1, color="black" )
          , panel.background = element_rect(fill = NA, color = "black")
          )
```

### Drivers
```{r}
brks <- c(0, 0.25, 0.5, 0.75, 1)

swerb_data %>% 
  filter(nchar(drivers) == 3) %>%
  filter((trip == "Morning" & event == "B") |  trip == "Afternoon" & event == "E") %>%
  filter(!is.na(observers)) %>%
  filter(driver != 'JKW') %>%
  group_by(driver, year_quarter, trip) %>%
  mutate(trips=n()) %>% 
  group_by(driver, year_quarter, trip, trips, adcode_description) %>%
  summarise(counts=n()) %>% 
  mutate(perc=counts/trips) %>%
  ggplot(aes(x = factor(year_quarter), y = perc, fill = factor(adcode_description))) +
    geom_bar(stat="identity", width = 0.7) +
    geom_text(aes(label=trips, factor(year_quarter), y = 0.1), size = 3, color = "black") +
    scale_y_continuous(breaks = brks, labels = scales::percent(brks)) +
    labs(x = "Year-month", y = "Percentages", fill = "Ascent and Descent codes") +
    scale_fill_manual(values=adcode_colors)  +
    facet_grid(trip~driver) +
    ggtitle("Proportion of adcodes") +
    ggthemes::theme_tufte() +
    theme(axis.text.x = element_text(angle = 90)
          , legend.position = 'bottom'
          , panel.grid.major.y = element_line( size=.1, color="black" )
          )
```

